STR: Seasonal-Trend decomposition using Regression

### 2020-11-17 str_ijds.tex
Copying relevant files from Alex's bitbucket private repo

---
title: "STR: Seasonal-Trend decomposition using Regression"
author:
 - familyname: Hyndman
   othernames: Rob J
   address: Monash University\newline Australia
   email: Rob.Hyndman@monash.edu
   correspondingauthor: true
 - familyname: Dokumentov
   othernames: Alexander
   address: Growing Data\newline Australia
abstract: "We propose a new method for decomposing seasonal data: STR (a Seasonal-Trend decomposition using Regression). Unlike other decomposition methods, STR allows for multiple seasonal and cyclic components, covariates, seasonal patterns that may have non-integer periods, and seasonality with complex topology. It can be used for time series with any regular time index including hourly, daily, weekly, monthly or quarterly data. It is competitive with existing methods when they exist, but tackles many more decomposition problem than other methods allow.\\par STR is based on a regularized optimization, and so is somewhat related to ridge regression. Because it is based on a statistical model, we can easily compute confidence intervals for components, something that is not possible with most existing decomposition methods (such as STL, X-12-ARIMA, SEATS-TRAMO, etc.).\\par
  Our model is implemented in the R package *stR*, so can be applied by anyone  to their own data."
keywords: "time series decomposition, seasonal data, Tikhonov regularisation, ridge regression, LASSO, STL, TBATS, X-12-ARIMA, BSM"
blind: false
cover: false
toc: false
bibliography: strrefs.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    toc: no
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
header-includes:
  - \graphicspath{{figs/}}
  - \def\addlinespace{}
  - \usepackage{amsmath}
  - \setlength{\arraycolsep}{3pt}
  - \setcounter{MaxMatrixCols}{20}
  - \newcommand{\argmin}{\operatorname{arg\,min}}
  - \newcommand{\sn}{\kappa}
  - \renewcommand{\vec}{\operatorname{vec}}
  - \newcommand{\diag}{\operatorname{diag}}
  - \renewcommand{\mod}{\operatorname{mod}}
  - \usepackage[section]{placeins}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  cache=TRUE
)
library(tidyverse)
library(stR)
library(fabletools)
library(distributional)
library(lubridate)

# Create predictor dist object
predictor_dist <- function(x, level) {
    dist_percentile(
      with(x, mapply(c, data, split(lower, row(lower)), split(upper, row(upper)), SIMPLIFY = FALSE)),
      list(c(50, 50 + rep(c(-1, 1), each = length(level))*rep(level, 2)/2))
    )
  }
```

# Time series decomposition

@@

1. It helps separate and understand trend and seasonal effects within a time series. This allows a greater understanding of peak times, and an estimation of the relative effect of holidays, working days and weekends. These can then be used in resource planning.
2. It allows the estimation of seasonally-adjusted data, which can then be used to assess the health of a business or industry after adjusting for periodic effects.  This is done, for example, with labour market statistics where the seasonally adjusted unemployment rate is used by governments in assessing the state of the economy.
3. It can be used to identify anomalies by eliminating the predictable variation due to trend and seasonality, leaving the unpredictable variation (the "remainder"). Any anomalies in the data are revealed in this "remainder" component, when they are often hidden in the original series. Anomaly detection helps organizations identify problems that may need attention. For example, many organizations measure millions of time series at high frequency and some anomaly detection is essential to identify where analysts should address their attention.

Existing time series decomposition methods are designed for monthly and quarterly data, with few tools available for more frequent data. The STL procedure of @cleveland1990stl is the only widely available decomposition tool for data observed more frequently than monthly, but it assumes a very simple additive structure and allows no covariates. This makes it unsuitable for a large number of problems.

@@

## A brief history of time series decomposition

The first attempts to decompose time series into various components can be dated to the 19th century, when @Poynting1884 proposed price averaging as a tool for eliminating trend and seasonal fluctuations. Later, his approach was extended by @hooker1901suspension, @spencer1904graduation and @anderson1914nochmals. Further research in this direction included that by @copeland1915statistical, who was the first to attempt to extract the seasonal component, until @macaulay1930smooth proposed a method which gradually became "classical". The work of Macaulay led to the Census II method [@shiskin1957electronic] which became popular amongst national statistics offices. The Census II method has continued to evolve, with additional features including robustness, calendar effects, covariates, ARIMA extensions, and extensive diagnostics. Widely used versions of this approach have been X-11 [@shishkin1967x], X-11-ARIMA [@dagum1988x11arima;@ladiray2001seasonal], X-12-ARIMA [@findley1998new] and X-13-ARIMA-SEATS [@findley2005some].

A different approach was followed by @cleveland1990stl, who developed STL (Seasonal-Trend decomposition using Loess), which has come to be widely used outside the national statistics agencies, partly because of its availability in R [@R]. This method uses iterative Loess smoothing to obtain an estimate of the trend, and then Loess smoothing again to extract a changing additive seasonal component.

Several model-based methods for seasonal decomposition have been developed, including the TRAMO/SEATS procedure which was developed at the Bank of Spain [@Gomez2001], the TBATS model of @de2011forecasting, and various structural time series model approaches [@harvey1990forecasting;@Commandeur2011]. One big advantage of using a model for seasonal decomposition and adjustment is that it provides a natural way to compute confidence and prediction intervals.

@@

  * provides a meaningful and simple statistical model;
  * allows for computation of confidence intervals of components;
  * can take account of covariates;
  * allows for fractional seasonal periods (e.g., with weekly data);
  * allows for multiple seasonal periods (e.g., with daily or sub-daily data);
  * allows for complex seasonal topology (e.g., due to public holidays);
  * can incorporate covariates which interact with the seasonality.

@@

# STR models

We suppose our time series $y_{t}$ can be decomposed additively as follows:
\begin{equation}
  \label{STR:many_parts}
  y_{t} = T_{t} + \sum_{i=1}^{I} S^{(i)}_{t} + \sum_{p=1}^P \phi_{p,t} z_{t,p} + R_{t} ,
\end{equation}
where

  * $T_{t}$ is a smoothly changing trend;
  * $S^{(i)}_{t}$ are smoothly changing seasonal components with possibly complex topology;
  * $z_{p,t}$ are covariates with coefficients $\phi_{p,t}$ which may be time-varying and even seasonal;
  * $R_{t}$ is the "remainder".

@@

Where components behave multiplicatively, we can use a log or Box-Cox transformation [@BC64] to made the additive assumption more reasonable.

@@

## Smoothness constraints via difference operators

Smoothness of a function implies that its second derivatives are small, since otherwise large values of the second derivatives would lead to high curvature of the function making it "wiggly" [see @Wood]. Thus, we consider a function smooth if its second derivatives are normally distributed with zero mean and where the variance controls the degree of smoothness.

Because time is discrete, we approximate derivatives with differences. Here, we use the notation $\Delta_2$ for the double difference operator, $\Delta_2 = (1-B)^2$, where $B$ is the backshift operator.

So a smooth trend is obtained by requiring that $\Delta_2 T_t = T_{t}-2T_{t-1}+T_{t-2}$ are iid $\mathcal{N}(0,\sigma_L^2)$, where $\sigma_L$ controls the degree of smoothness. This can be expressed via the multivariate normal density
$$
  f(\bm{D}_\ell \bm{\ell}) = f_L(\sigma_L) \exp\left\{-\frac{1}{2}\big\|\bm{D}_\ell \bm{\ell} / \sigma_L\big\|_{L_2}^2\right\},
$$
where
$\bm{\ell} = \langle T_{t} \rangle_{t=1}^{n}$ is an $n$-vector of the trend component;
$\bm{D}_\ell$ is the $(n-2) \times n$ difference operator matrix such that $\bm{D}_\ell\bm{\ell} = \langle\Delta^2 T_{t}\rangle_{t=3}^n$;
and $f_L(\sigma_L)$ is a normalizing function.

## Smooth two-dimensional seasonal surfaces

Let $m_i$ denote the number of "seasons" in the seasonal component $S^{(i)}_{t}$. For example, if $y_t$ is monthly data, there is only one seasonal component ($I=1$) with $m_1 = 12$; if $y_t$ is daily data, there may be two seasonal components ($I=2$), with a weekly pattern ($m_1=7$) and an annual pattern ($m_2 = 365$ ignoring leap years). Let us also define the function $\kappa_i(t) = t\mod m_i$, which transforms time $t$ into the corresponding season $\kappa_i(t) \in \{1,\dots,m_i\}$.

At time $t$, we observe only one element of each seasonal component. It is reasonable to ask what the other elements of the component are at that moment $t$. For example, if we have daily data with weekly seasonality, it is valid to ask on a Wednesday, "What is the current value of the seasonal component corresponding to Friday?". In other words, we define latent components that are responsible for seasons other than $\kappa_i(t)$. In this way, we treat the $i$th seasonal component as a two-dimensional matrix, $[S^{(i)}_{k,t}]$ ($k=1,\dots,m_i;t=1,\dots,n$). This formulation of each seasonal component as a two-dimensional structure is one of the novelties and features of STR, and it allows for simple and natural smoothing constraints to be applied.

The seasonal two-dimensional surface actually has the topology of a cylinder, where dimension $t$ (time) is along the length of the cylinder, but dimension $s$ (season) is "circular" around the cylinder. For example, in the case of a weekly seasonal component in daily data, the season dimension corresponds to days of the week. A time series corresponds to a spiral on the cylinder.  The other locations on the cylinder (between the spirals) will represent "imaginary" seasonality. These correspond, for example, to the current value of Friday's seasonality when today is a Wednesday.

To ensure they remain identifiable and interpretable, the seasonal terms must sum to zero at any time $t$, so they must satisfy the property $\sum\limits_k S^{(i)}_{k,t} = 0$ for each $t$.

Smoothness of the weekly seasonal pattern can be described as smoothness of $S^{(i)}_{k,t}$ in the seasonal direction $k$, while the speed at which the seasonality changes over time is controlled by the smoothness of $S^{(i)}_{k,t}$ in the time direction $t$. Moreover, some other features are possible to express. For example, by restricting second partial derivatives of $S^{(i)}_{k,t}$ in both time and season directions, it is possible to describe the weekly seasonal pattern changing "synchronously" in the time-season direction.

The seasonal smoothness can also be expressed using difference operator matrices. We define three such matrices, $\bm{D}_{tt,i}$, $\bm{D}_{ss,i}$ and $\bm{D}_{st,i}$, corresponding to smoothness in the time, season and time-season directions respectively.

For smoothness in the time direction, we require that the seasonal vectors $\langle \Delta_2 S^{(i)}_{k,t} \rangle_{k=1}^{m_i}$ are iid $\mathcal{N}(0,\sigma_{i}^2 \bm{\Sigma}_{i} )$, where $\sigma_i^2$ controls the degree of smoothness and $\bm{\Sigma}_{i} = \bm{I}_{m_i} - \frac1m\bm{1}_{m_i}$ is the covariance matrix of the $m_i$ random variables $\xi_k = \eta_k - m_i^{-1}\sum\limits_{r = 1}^{m_i}\eta_r$, obtained from the iid standard normal random variables $\eta_1,\dots,\eta_{m_i}$.

Let $\bm{S}^{(i)} = [S_{k,t}^{(i)}]$ contain the $i$th seasonal surface, $\bm{S}_i^{-}$ be the matrix $\bm{S}^{(i)}$ without the last row, and let $\bm{s}_i = \vec(\bm{S}_i^{-})$ be a vector of length $n(m_i-1)$ representing the $i$th seasonal component in vector form. Thus we can write the second differences of the seasonal component along the time dimension as  $\bm{D}_{tt,i} \bm{s}_i  = \langle \Delta^2_{t} \bm{S}^{(i)}_{k,t} \rangle_{k=1}^{m_i}$, $3 \le t \le n$, for a suitably defined $(n-2)m_i \times nm_i$ difference operator matrix $\bm{D}_{tt,i}$.

>>>>

\documentclass[11pt,a4paper,]{article}
\usepackage{lmodern}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={STR: Seasonal-Trend decomposition using Regression},
            pdfkeywords={time series decomposition, seasonal data, Tikhonov regularisation, ridge regression, LASSO, STL, TBATS, X-12-ARIMA, BSM},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{strrefs.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\title{STR: Seasonal-Trend decomposition using Regression}

%% MONASH STUFF

%% CAPTIONS
\RequirePackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}

%% FONT
\RequirePackage{bera}
\RequirePackage{mathpazo}

%% HEADERS AND FOOTERS
\RequirePackage{fancyhdr}
\pagestyle{fancy}
\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
\makeatletter
\lhead{\textsf{\expandafter{\@title}}}
\makeatother
\rhead{}
\cfoot{}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{\sffamily\thepage} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%% MATHS
\RequirePackage{bm,amsmath}
\allowdisplaybreaks

%% GRAPHICS
\RequirePackage{graphicx}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.8}

%\RequirePackage[section]{placeins}

%% SECTION TITLES
\RequirePackage[compact,sf,bf]{titlesec}
\titleformat{\section}[block]
  {\fontsize{15}{17}\bfseries\sffamily}
  {\thesection}
  {0.4em}{}
\titleformat{\subsection}[block]
  {\fontsize{12}{14}\bfseries\sffamily}
  {\thesubsection}
  {0.4em}{}
\titlespacing{\section}{0pt}{*5}{*1}
\titlespacing{\subsection}{0pt}{*2}{*0.2}

%% TITLE PAGE
\def\Date{\number\day}
\def\Month{\ifcase\month\or
 January\or February\or March\or April\or May\or June\or
 July\or August\or September\or October\or November\or December\fi}
\def\Year{\number\year}

\makeatletter
\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
\def\nojel{\def\showjel{}}
\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
\def\cover{{\sffamily\setcounter{page}{0}
        \thispagestyle{empty}
        \placefig{2}{1.5}{width=5cm}{monash2}
        \placefig{16.9}{1.5}{width=2.1cm}{MBusSchool}
        \begin{textblock}{4}(16.9,4)ISSN 1440-771X\end{textblock}
        \begin{textblock}{7}(12.7,27.9)\hfill
        \includegraphics[height=0.7cm]{AACSB}~~~
        \includegraphics[height=0.7cm]{EQUIS}~~~
        \includegraphics[height=0.7cm]{AMBA}
        \end{textblock}
        \vspace*{2cm}
        \begin{center}\Large
        Department of Econometrics and Business Statistics\\[.5cm]
        \footnotesize http://monash.edu/business/ebs/research/publications
        \end{center}\vspace{2cm}
        \begin{center}
        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
                \LARGE\@author\end{onehalfspace}
        }}
        \end{center}
        \vfill
                \begin{center}\Large
                \Month~\Year\\[1cm]
                Working Paper \@wp
        \end{center}\vspace*{2cm}}}
\def\pageone{{\sffamily\setstretch{1}%
        \thispagestyle{empty}%
        \vbox to \textheight{%
        \raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
        \vspace{2cm}\par
        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
        \hspace{1cm}\showjel\vss}}}
\def\blindtitle{{\sffamily
     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
        }}
\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}

\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
\let\maketitle\titlepage
\makeatother

%% SPACING
\RequirePackage{setspace}
\spacing{1.5}

%% LINE AND PAGE BREAKING
\sloppy
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\RequirePackage{microtype}

%% PARAGRAPH BREAKS
\setlength{\parskip}{1.4ex}
\setlength{\parindent}{0em}

%% HYPERLINKS
\RequirePackage{xcolor} % Needed for links
\definecolor{darkblue}{rgb}{0,0,.6}
\RequirePackage{url}

\makeatletter
\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
\makeatother
\hypersetup{
     citecolor=0 0 0,
     breaklinks=true,
     bookmarksopen=true,
     bookmarksnumbered=true,
     linkcolor=darkblue,
     urlcolor=blue,
     citecolor=darkblue,
     colorlinks=true}

%% KEYWORDS
\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}

%% ABSTRACT
\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
  {\end{minipage}}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage[showonlyrefs]{mathtools}
\usepackage[no-weekday]{eukdate}

%% BIBLIOGRAPHY

\makeatletter
\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
\makeatother
\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false}

\DeclareFieldFormat{url}{\texttt{\url{#1}}}
\DeclareFieldFormat[article]{pages}{#1}
\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
\DeclareFieldFormat[inproceedings]{title}{#1}
\DeclareFieldFormat{shorthandwidth}{#1}
% No dot before number of articles
\usepackage{xpatch}
\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
% Remove In: for an article.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}

\makeatletter
\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
\makeatother
\renewcommand*{\finalnamedelim}{%
  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
  \addspace\&\space}

\nojel

\RequirePackage[absolute,overlay]{textpos}
\setlength{\TPHorizModule}{1cm}
\setlength{\TPVertModule}{1cm}
\def\placefig#1#2#3#4{\begin{textblock}{.1}(#1,#2)\rlap{\includegraphics[#3]{#4}}\end{textblock}}

\nocover

\author{Rob J~Hyndman, Alexander~Dokumentov}
\addresses{\textbf{Rob J Hyndman}\newline
Monash University\newline Australia
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\newline Corresponding author\\[1cm]
\textbf{Alexander Dokumentov}\newline
Growing Data\newline Australia
\\[1cm]
}

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf Hyndman, Dokumentov: \@date}
\makeatother

\graphicspath{{figs/}}
\def\addlinespace{}
\usepackage{amsmath}
\setlength{\arraycolsep}{3pt}
\setcounter{MaxMatrixCols}{20}
\newcommand{\argmin}{\operatorname{arg\,min}}
\newcommand{\sn}{\kappa}
\renewcommand{\vec}{\operatorname{vec}}
\newcommand{\diag}{\operatorname{diag}}
\renewcommand{\mod}{\operatorname{mod}}
\usepackage[section]{placeins}

\begin{document}
\maketitle
\begin{abstract}
We propose a new method for decomposing seasonal data: STR (a Seasonal-Trend decomposition using Regression). Unlike other decomposition methods, STR allows for multiple seasonal and cyclic components, covariates, seasonal patterns that may have non-integer periods, and seasonality with complex topology. It can be used for time series with any regular time index including hourly, daily, weekly, monthly or quarterly data. It is competitive with existing methods when they exist, but tackles many more decomposition problem than other methods allow.

\par

STR is based on a regularized optimization, and so is somewhat related to ridge regression. Because it is based on a statistical model, we can easily compute confidence intervals for components, something that is not possible with most existing decomposition methods (such as STL, X-12-ARIMA, SEATS-TRAMO, etc.).

\par

Our model is implemented in the R package \emph{stR}, so can be applied by anyone to their own data.
\end{abstract}
\begin{keywords}
time series decomposition, seasonal data, Tikhonov regularisation, ridge regression, LASSO, STL, TBATS, X-12-ARIMA, BSM
\end{keywords}

\hypertarget{time-series-decomposition}{%
\section{Time series decomposition}\label{time-series-decomposition}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  It helps separate and understand trend and seasonal effects within a time series. This allows a greater understanding of peak times, and an estimation of the relative effect of holidays, working days and weekends. These can then be used in resource planning.
\item
  It allows the estimation of seasonally-adjusted data, which can then be used to assess the health of a business or industry after adjusting for periodic effects. This is done, for example, with labour market statistics where the seasonally adjusted unemployment rate is used by governments in assessing the state of the economy.
\item
  It can be used to identify anomalies by eliminating the predictable variation due to trend and seasonality, leaving the unpredictable variation (the ``remainder''). Any anomalies in the data are revealed in this ``remainder'' component, when they are often hidden in the original series. Anomaly detection helps organizations identify problems that may need attention. For example, many organizations measure millions of time series at high frequency and some anomaly detection is essential to identify where analysts should address their attention.
\end{enumerate}

Existing time series decomposition methods are designed for monthly and quarterly data, with few tools available for more frequent data. The STL procedure of \textcite{cleveland1990stl} is the only widely available decomposition tool for data observed more frequently than monthly, but it assumes a very simple additive structure and allows no covariates. This makes it unsuitable for a large number of problems.

@@

\hypertarget{a-brief-history-of-time-series-decomposition}{%
\subsection{A brief history of time series decomposition}\label{a-brief-history-of-time-series-decomposition}}

The first attempts to decompose time series into various components can be dated to the 19th century, when \textcite{Poynting1884} proposed price averaging as a tool for eliminating trend and seasonal fluctuations. Later, his approach was extended by \textcite{hooker1901suspension}, \textcite{spencer1904graduation} and \textcite{anderson1914nochmals}. Further research in this direction included that by \textcite{copeland1915statistical}, who was the first to attempt to extract the seasonal component, until \textcite{macaulay1930smooth} proposed a method which gradually became ``classical''. The work of Macaulay led to the Census II method \autocite{shiskin1957electronic} which became popular amongst national statistics offices. The Census II method has continued to evolve, with additional features including robustness, calendar effects, covariates, ARIMA extensions, and extensive diagnostics. Widely used versions of this approach have been X-11 \autocite{shishkin1967x}, X-11-ARIMA \autocite{dagum1988x11arima,ladiray2001seasonal}, X-12-ARIMA \autocite{findley1998new} and X-13-ARIMA-SEATS \autocite{findley2005some}.

A different approach was followed by \textcite{cleveland1990stl}, who developed STL (Seasonal-Trend decomposition using Loess), which has come to be widely used outside the national statistics agencies, partly because of its availability in R \autocite{R}. This method uses iterative Loess smoothing to obtain an estimate of the trend, and then Loess smoothing again to extract a changing additive seasonal component.

Several model-based methods for seasonal decomposition have been developed, including the TRAMO/SEATS procedure which was developed at the Bank of Spain \autocite{Gomez2001}, the TBATS model of \textcite{de2011forecasting}, and various structural time series model approaches \autocite{harvey1990forecasting,Commandeur2011}. One big advantage of using a model for seasonal decomposition and adjustment is that it provides a natural way to compute confidence and prediction intervals.

@@

\begin{itemize}
\tightlist
\item
  provides a meaningful and simple statistical model;
\item
  allows for computation of confidence intervals of components;
\item
  can take account of covariates;
\item
  allows for fractional seasonal periods (e.g., with weekly data);
\item
  allows for multiple seasonal periods (e.g., with daily or sub-daily data);
\item
  allows for complex seasonal topology (e.g., due to public holidays);
\item
  can incorporate covariates which interact with the seasonality.
\end{itemize}

@@

\hypertarget{str-models}{%
\section{STR models}\label{str-models}}

We suppose our time series \(y_{t}\) can be decomposed additively as follows:
\begin{equation}
  \label{STR:many_parts}
  y_{t} = T_{t} + \sum_{i=1}^{I} S^{(i)}_{t} + \sum_{p=1}^P \phi_{p,t} z_{t,p} + R_{t} ,
\end{equation}
where

\begin{itemize}
\tightlist
\item
  \(T_{t}\) is a smoothly changing trend;
\item
  \(S^{(i)}_{t}\) are smoothly changing seasonal components with possibly complex topology;
\item
  \(z_{p,t}\) are covariates with coefficients \(\phi_{p,t}\) which may be time-varying and even seasonal;
\item
  \(R_{t}\) is the ``remainder''.
\end{itemize}

@@

Where components behave multiplicatively, we can use a log or Box-Cox transformation \autocite{BC64} to made the additive assumption more reasonable.

@@

\hypertarget{smoothness-constraints-via-difference-operators}{%
\subsection{Smoothness constraints via difference operators}\label{smoothness-constraints-via-difference-operators}}

Smoothness of a function implies that its second derivatives are small, since otherwise large values of the second derivatives would lead to high curvature of the function making it ``wiggly'' \autocite[see][]{Wood}. Thus, we consider a function smooth if its second derivatives are normally distributed with zero mean and where the variance controls the degree of smoothness.

Because time is discrete, we approximate derivatives with differences. Here, we use the notation \(\Delta_2\) for the double difference operator, \(\Delta_2 = (1-B)^2\), where \(B\) is the backshift operator.

So a smooth trend is obtained by requiring that \(\Delta_2 T_t = T_{t}-2T_{t-1}+T_{t-2}\) are iid \(\mathcal{N}(0,\sigma_L^2)\), where \(\sigma_L\) controls the degree of smoothness. This can be expressed via the multivariate normal density
\[
  f(\bm{D}_\ell \bm{\ell}) = f_L(\sigma_L) \exp\left\{-\frac{1}{2}\big\|\bm{D}_\ell \bm{\ell} / \sigma_L\big\|_{L_2}^2\right\},
\]
where
\(\bm{\ell} = \langle T_{t} \rangle_{t=1}^{n}\) is an \(n\)-vector of the trend component;
\(\bm{D}_\ell\) is the \((n-2) \times n\) difference operator matrix such that \(\bm{D}_\ell\bm{\ell} = \langle\Delta^2 T_{t}\rangle_{t=3}^n\);
and \(f_L(\sigma_L)\) is a normalizing function.

\hypertarget{smooth-two-dimensional-seasonal-surfaces}{%
\subsection{Smooth two-dimensional seasonal surfaces}\label{smooth-two-dimensional-seasonal-surfaces}}

Let \(m_i\) denote the number of ``seasons'' in the seasonal component \(S^{(i)}_{t}\). For example, if \(y_t\) is monthly data, there is only one seasonal component (\(I=1\)) with \(m_1 = 12\); if \(y_t\) is daily data, there may be two seasonal components (\(I=2\)), with a weekly pattern (\(m_1=7\)) and an annual pattern (\(m_2 = 365\) ignoring leap years). Let us also define the function \(\kappa_i(t) = t\operatorname{mod}m_i\), which transforms time \(t\) into the corresponding season \(\kappa_i(t) \in \{1,\dots,m_i\}\).

At time \(t\), we observe only one element of each seasonal component. It is reasonable to ask what the other elements of the component are at that moment \(t\). For example, if we have daily data with weekly seasonality, it is valid to ask on a Wednesday, ``What is the current value of the seasonal component corresponding to Friday?''. In other words, we define latent components that are responsible for seasons other than \(\kappa_i(t)\). In this way, we treat the \(i\)th seasonal component as a two-dimensional matrix, \([S^{(i)}_{k,t}]\) (\(k=1,\dots,m_i;t=1,\dots,n\)). This formulation of each seasonal component as a two-dimensional structure is one of the novelties and features of STR, and it allows for simple and natural smoothing constraints to be applied.

The seasonal two-dimensional surface actually has the topology of a cylinder, where dimension \(t\) (time) is along the length of the cylinder, but dimension \(s\) (season) is ``circular'' around the cylinder. For example, in the case of a weekly seasonal component in daily data, the season dimension corresponds to days of the week. A time series corresponds to a spiral on the cylinder. The other locations on the cylinder (between the spirals) will represent ``imaginary'' seasonality. These correspond, for example, to the current value of Friday's seasonality when today is a Wednesday.

To ensure they remain identifiable and interpretable, the seasonal terms must sum to zero at any time \(t\), so they must satisfy the property \(\sum\limits_k S^{(i)}_{k,t} = 0\) for each \(t\).

Smoothness of the weekly seasonal pattern can be described as smoothness of \(S^{(i)}_{k,t}\) in the seasonal direction \(k\), while the speed at which the seasonality changes over time is controlled by the smoothness of \(S^{(i)}_{k,t}\) in the time direction \(t\). Moreover, some other features are possible to express. For example, by restricting second partial derivatives of \(S^{(i)}_{k,t}\) in both time and season directions, it is possible to describe the weekly seasonal pattern changing ``synchronously'' in the time-season direction.

The seasonal smoothness can also be expressed using difference operator matrices. We define three such matrices, \(\bm{D}_{tt,i}\), \(\bm{D}_{ss,i}\) and \(\bm{D}_{st,i}\), corresponding to smoothness in the time, season and time-season directions respectively.

For smoothness in the time direction, we require that the seasonal vectors \(\langle \Delta_2 S^{(i)}_{k,t} \rangle_{k=1}^{m_i}\) are iid \(\mathcal{N}(0,\sigma_{i}^2 \bm{\Sigma}_{i} )\), where \(\sigma_i^2\) controls the degree of smoothness and \(\bm{\Sigma}_{i} = \bm{I}_{m_i} - \frac1m\bm{1}_{m_i}\) is the covariance matrix of the \(m_i\) random variables \(\xi_k = \eta_k - m_i^{-1}\sum\limits_{r = 1}^{m_i}\eta_r\), obtained from the iid standard normal random variables \(\eta_1,\dots,\eta_{m_i}\).

Let \(\bm{S}^{(i)} = [S_{k,t}^{(i)}]\) contain the \(i\)th seasonal surface, \(\bm{S}_i^{-}\) be the matrix \(\bm{S}^{(i)}\) without the last row, and let \(\bm{s}_i = \operatorname{vec}(\bm{S}_i^{-})\) be a vector of length \(n(m_i-1)\) representing the \(i\)th seasonal component in vector form. Thus we can write the second differences of the seasonal component along the time dimension as \(\bm{D}_{tt,i} \bm{s}_i = \langle \Delta^2_{t} \bm{S}^{(i)}_{k,t} \rangle_{k=1}^{m_i}\), \(3 \le t \le n\), for a suitably defined \((n-2)m_i \times nm_i\) difference operator matrix \(\bm{D}_{tt,i}\).
